Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists

### 2024-04-25 paper.Rmd
updates

While Agent-Based Modelling importance in archaeological practice is steadily growing \_\_\_ !!!please add the following reference here: <https://www.researchgate.net/publication/376410267_Bibliometric_Analysis_of_Agent-Based_Simulation_in_Archaeology_People_Topics_and_Future_Prospects> \_\_\_, there are significant bariers to its wider adoption. One of the sectorial problems is the access to vocational training required to gain the necessary computer programming and modelling skills. A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support i.e. 27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school while 14% received training as part of a degree.

@@

[!!!COMMENT: this sentence should go into the acknowledgements] This project, Agent-Based Modelling for Archaeologists (<https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>), was delivered with the support of the Erasmus+ programme of the European Union.

@@

-   Development of the framework for skills acquisition aligned with the Digital Skills Passport !!!\_\_ add ref here!

@@

3.  !!! [Comment: This is very unclear. Currently it reads as if taking the course will make people emigrage to other countries. I don't think that is in any way true. Also, most readers would not know what European or Sectoral Framework we are referring to] improved transnational mobility through matching of the skills being delivered with European and Sectoral Frameworks, leading to individual workers being better enabled to demonstrate their skills to employers across Europe;

@@

| Conference or event          | Period                  | Online, in person or hybrid | Number of participants | Number of registrations |
|---------------|---------------|---------------|---------------|---------------|
| CAA Amsterdam                | April 2023              | In person                   | 36                     | 53                      |
| EAA Belfast                  | August 2023             | In person                   | 21                     | No registration         |
| CAA-DE/NL-Fl Online workshop | October 2023            | Online                      | 40                     | 76                      |
| Reuvensdagen                 | November 2023           | In person                   | 13                     | No registration         |
| CAA-UK                       | November 2023           | Hybrid                      | 53                     |                         |
| Leiden (course)              | December 2023           | In person                   | 11                     | No registration         |
| Aarhus                       | January/February 2024   | Online                      | 176                    | \>500                   |
| Saxion                       | March 2023-January 2024 | In person                   | 18                     | No registration         |
| **Total**                    |                         |                             | **368**                | **\>628**               |

: Events at which the tutorials were delivered and tested.

@@

Over the course of the project three groups of students worked on the project during the Smart Solutions Semester at Saxion University of Applied Sciences. This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (<https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions>). The backgrounds of the various students were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media & Game Technologies, and ICT. The diverse groups of students contributed new ideas, developing educational material, testing the tutorials and developing a style for the website and other materials. !!! COMMENT- this section needs to be more specific. Why did these student work on the project? At what capacity? What did they contribute? etc

@@

## ABM tutorials

@@

In tutorial 2 the user learns the basics of programming in NetLogo and develops their first simulation based on a previously published model of the Out of Africa dispersal of *Homo sapiens !!! Pls insert the following reference*

Young, David A., and Robert L. Bettinger. 1992. “The Numic Spread: A Computer Simulation.” *American Antiquity* 57 (1): 85–99. <https://doi.org/10.2307/2694836>.

. The learner works through the basics of NetLogo syntax and learn how to set up a simulation and visualize its outcomes. This tutorial consists of 9 lessons. This tutorial intends to guide the learner from the introduction to the beginner level of proficiency. The learning curve is relatively gentle. This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed. The tutorial intoduces several most commonly used built-in functions in Netlogo (called *primitives*) are introduces the general structure of all models cosisting of the initialization phase and the main simulation loop. It walks the participants through te most important NetLogo settings, such as dimensions, coordinates and the origin point of the grid on which the simulation runs. The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements. The next level is achieved by introducing the topic of function definition and different types of variables. Exporting data produced by the model through plots is also explained.

In the third tutorial, the learner builds a simple trade model, again based on a published simulation !!! pls insert the following reference:

Romanowska, Iza. 2018. “‪Using Agent-Based Modelling to Infer Economic Processes in the Past‬.” In *‪Cuantificar Las Economías Antiguas: Problemas y Métodos*, Publicacions i Edicions de la Universitat de Barcelona, 107–18. Barcelona: Universitat de Barcelona. <https://scholar.google.com/citations?view_op=view_citation&hl=en&user=tdcyWfoAAAAJ&citation_for_view=tdcyWfoAAAAJ:YOwf2qJgpHMC>.

Here learners are expanding their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters. The learner is introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity. This tutorial consists of 7 lessons. This tutorial intends to guide the learner from the beginner to the intermediate level of proficiency. This is achieved by introducing standards of code development such as modular code, pseudocode, debugging and annotating the code. In addition, the tutorial introduces custom agent breeds, visualization with labels and reporters, plots and monitors. This takes the learner to a more advanced level, since it develops knowledge of the programming language combined with diagnosing unexpected behaviour and problem solving.

In tutorial 4 the learner works with a seminal Sugarscape model !!! pls insert the following ref.

Epstein, Joshua M., and Robert L. Axtell. 1996. *Growing Artificial Societies. Social Science from the Bottom Up*. Boston: MIT Press.

to further expand their agent-based modelling and NetLogo skills. The learner explores how to set up more complex interactions between agents and the environment. Furthermore, the principles of setting up good experiments and validating models is explained. This tutorial consists of 8 lessons. This tutorial intends to solidify the learner's understanding at the intermediate level of proficiency by introducing yet more complex agent-environment interaction, alternative ways to visualize the environment and modelling spatial dynamics. Creating toy landscapes is central to lesson 5. The learner dives into the topic of experiment design at a practical level but also with regards to the scientific principles, such as validation. Thus they are guided through the process of setting up experiments in NetLogo and collecting results using monitors and plots. The topic of validation of agent-based models is critical for most archaelogical applications and thus several ways to compare simulation results with the archaeological record are explored.

@@

A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants). The respondents came from !!! COMMENT PLS ADD THE NUMBER countries that represented almost all continents (see Figure \@ref(fig:nationality). There was a clear skew towards participants from !!! COMMENT what part of the world? Europe? South America. The number of women and men in our sample were close to equal (see Figure \@ref(fig:gender-age)). The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years old indicating that the audience was predominantly composed of early career archaeologists. It seems that the different events also had a slightly different distribution of both gender and age. For example, more older people attended the workshop at the CAA conference in April 2023 and more men were present during the workshop at the Reuvensdagen in November 2023. COMMENT - in the last sentence, are these significant differences? And if so, please provide the possible reasons for these differences.

@@

Most of the respondents reported having a level of computer literacy already including particular computational skills common in archaeology, for example !!! COMMENT pls add the most popular ones (see Figure \@ref(fig:computer-skills)), and !!! COMMENT PLS insert the % number here had some prior knowledge of ABM, while !!! COMMENT PLS insert the % number here was completely new to the subject. The majority !!! COMMENT PLS insert the % number here had never applied ABM to their research before participating in the workshops with only 14 participants who had experience of developing archaeological ABM. It is interesting to note that many respondents (!!! COMMENT PLS insert the % number here) did know what kind of software was available for ABM (see Figure \@ref(fig:abm-knowledge)).

@@

As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before. The respondents were also asked how they rated the availability of ABM educational resources (Figure \@ref(fig:available-theory)). While *!!! COMMENT PLS insert the % number here* of them had no opinion on the subject, a !!! COMMENT PLS insert the % number here proportion of the responded answered that they rated the available theory on ABM as limited and only a smaller group !!! COMMENT PLS insert the % number here as sufficient or as more than sufficient. This clearly shows the need for more and better educational material.

@@

*!!! COMMENT PLS insert the % number here* of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good. The teachers were rated even higher than the teaching material, with more respondents giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).

@@

Two open questions were asked to the respondents: *!!! COMMENT PLS spell out the questions*

@@

We also asked the participants about their intentions in terms of ABM and their professional life. The majority of the respondents - 88%, wanted to apply ABM in the future in their research (Figure \@ref(fig:future-abm)). Some of the respondents, *!!! COMMENT PLS insert the % number here,* were already using ABM in their work. A large group *!!! COMMENT PLS insert the % number here* was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities. Many respondents also shared the context to which they were thinking to apply ABM mentioning topics such as movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms. Others thought of demography, social networks, migration or settlement distributions patterns. The natural environment and the interaction with humans in the past was also mentioned frequently, often in relation to GIS and the question of how complimentary these two computational techniques are. The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.

@@

As mentioned above, the following competence area's were identified as relevant for the tutorials: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving [@europeancommissionjointresearchcentre2022]. In these competence area's various competences are developed towards an advanced or specialised level.

@@

Funding acknowledgements: the publication was financed by The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382); IR was supported by the European Union’s Horizon 2020 Research and
Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513 and the
Aarhus University Research Foundation.

>>>>

While Agent-Based Modelling importance in archaeological practice is steadily growing [@romanowska2023] there are significant bariers to its wider adoption. One of the sectorial problems is the access to vocational training required to gain the necessary computer programming and modelling skills. A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support i.e. 27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school while only 14% received training as part of a degree.

@@

This project, Agent-Based Modelling for Archaeologists, was delivered with the support of the Erasmus+ programme of the European Union.

@@

-   Development of the framework for skills acquisition aligned with the Digital Skills Passport [DOUG: reference]?

@@

3.  improved transnational mobility through matching of the skills being delivered with European and Sectoral Frameworks for education, leading to individual workers being better enabled to demonstrate their skills to employers across Europe;

@@

| Conference or event          | Period                  | Online, in person or hybrid | Number of participants | Number of registrations |
|------------------------------|-------------------------|-----------------------------|------------------------|-------------------------|
| CAA Amsterdam                | April 2023              | In person                   | 36                     | 53                      |
| EAA Belfast                  | August 2023             | In person                   | 21                     | No registration         |
| CAA-DE/NL-Fl Online workshop | October 2023            | Online                      | 40                     | 76                      |
| Reuvensdagen                 | November 2023           | In person                   | 13                     | No registration         |
| CAA-UK                       | November 2023           | Hybrid                      | 53                     |                         |
| Leiden (course)              | December 2023           | In person                   | 11                     | No registration         |
| Aarhus                       | January/February 2024   | Online                      | 176                    | \>500                   |
| Saxion                       | March 2023-January 2024 | In person                   | 18                     | No registration         |
| **Total**                    |                         |                             | **368**                | **\>628**               |

: Conferences, events and situations were workshops were held at which the tutorials were tested.

@@

Over the course of the project three groups of students worked on the project during the Smart Solutions Semester at Saxion University of Applied Sciences. This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (<https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions>). The backgrounds of the various students were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media & Game Technologies, and ICT. The diverse groups of students contributed new ideas, developing educational material, testing the tutorials and developing a style for the website and other materials.

@@

## Final tutorials

@@

In tutorial 2 the user will learn the basics of NetLogo by making your their simulation on the Out of Africal dispersal of *Homo sapiens* [@young1995].The learner works through the basics of NetLogo syntax and learn how to set up a simulation and visualize its outcomes. This tutorial consists of 9 lessons. This tutorial intends to guide the learner from the introduction to the beginner level of proficiency. The learning curve is relatively gentle. This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed. The tutorial intoduces several most commonly used built-in functions in Netlogo (called *primitives*) are introduces the general structure of all models cosisting of the initialization phase and the main simulation loop. It walks the participants through te most important NetLogo settings, such as dimensions, coordinates and the origin point of the grid on which the simulation runs. The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements. The next level is achieved by introducing the topic of function definition and different types of variables. Exporting data produced by the model through plots is also explained.

In the third tutorial, the learner builds a simple trade model, again based on a published simulation [@romanowska2018]. Here learners are expanding their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters. The learner is introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity. This tutorial consists of 7 lessons. This tutorial intends to guide the learner from the beginner to the intermediate level of proficiency. This is achieved by introducing standards of code development such as modular code, pseudocode, debugging and annotating the code. In addition, the tutorial introduces custom agent breeds, visualization with labels and reporters, plots and monitors. This takes the learner to a more advanced level, since it develops knowledge of the programming language combined with diagnosing unexpected behaviour and problem solving.

In tutorial 4 the learner works with a seminal Sugarscape model [@epstein1996] to further expand their agent-based modelling and NetLogo skills. The learner explores how to set up more complex interactions between agents and the environment. Furthermore, the principles of setting up good experiments and validating models is explained. This tutorial consists of 8 lessons. This tutorial intends to solidify the learner's understanding at the intermediate level of proficiency by introducing yet more complex agent-environment interaction, alternative ways to visualize the environment and modelling spatial dynamics. Creating toy landscapes is central to lesson 5. The learner dives into the topic of experiment design at a practical level but also with regards to the scientific principles, such as validation. Thus they are guided through the process of setting up experiments in NetLogo and collecting results using monitors and plots. The topic of validation of agent-based models is critical for most archaelogical applications and thus several ways to compare simulation results with the archaeological record are explored.

@@

A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants). The respondents came from at least 40 countries that represented almost all continents (see Figure \@ref(fig:nationality). There was a clear skew towards participants from Europe. The number of female respondents slightly outnumbers the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)). The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years old indicating that the audience was predominantly composed of early career archaeologists. It seems that the different events also had a slightly different distribution of both gender and age. For example, more older people attended the workshop at the CAA conference in April 2023 and more men were present during the workshop at the Reuvensdagen in November 2023. This might be due the the differences of audiences at the conferences. For the Reuvensdagen the number of participants of the workshop was relatively low and this variance might be due to chance.

@@

Most of the respondents reported having a level of computer literacy already including particular computational skills common in archaeology, for example working with word processors, GIS and spreadsheets (see Figure \@ref(fig:computer-skills)), and 111 of the respondents had some prior knowledge of ABM, while 59 were completely new to the subject. The majority (158) had never applied ABM to their research before participating in the workshops with only 14 participants who had experience of developing archaeological ABM. It is interesting to note that many respondents (108) did know what kind of software was available for ABM (see Figure \@ref(fig:abm-knowledge)).

@@

As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before. The respondents were also asked how they rated the available knowledge on ABM (Figure \@ref(fig:available-theory)). While 64 of them had no opinion on the subject, a proportion of the respondents (28) answered that they rated the available theory on ABM as limited and only a small group (39) as sufficient or better. This clearly shows the need for more and better educational material.

@@

of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good. The teachers were rated even higher than the teaching material, with more respondents giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).

The majority of the respondents (153) were enthusiastic about the teaching material, with the majority rating it as excellent or good. The teachers were even rated better than the teaching material, with more respondents (165) giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).

@@

Two more open questions were asked to the respondents. The first was aimed at learning what aspects the participants liked in the tutorials (What aspects did you like most?), and the other was aimed at getting feedback to improve the tutorials (What would you like to see changed or what could be better/different?).

@@

We also asked the participants about their intentions in terms of ABM and their professional life. The majority of the respondents - 88%, wanted to apply ABM in the future in their research (Figure \@ref(fig:future-abm)). Some of the respondents were already using ABM in their work. A large group (around 20-30%) was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities. Many respondents also shared the context to which they were thinking to apply ABM mentioning topics such as movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms. Others thought of demography, social networks, migration or settlement distributions patterns. The natural environment and the interaction with humans in the past was also mentioned frequently, often in relation to GIS and the question of how complimentary these two computational techniques are. The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.

@@

As mentioned above, the following competence area's were identified as relevant for the tutorials: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving [@europeancommissionjointresearchcentre2022]. In these competence area's various competences are developed towards an advanced or even specialised level. A more extensive description of the Digital Competence Framework for Citizens in relation to Agent-Based Modelling for archaeologists is also available [@visser2024].

@@

Funding acknowledgements: the publication was financed by The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513 and the Aarhus University Research Foundation.
